---
title: "Love and Mazer - Phenological Responses to Climate"
author: "Natalie Love"
date: "1/1/2021"
output: html_document
---
This code was generated using R version 4.0.3 (Bunny-Wunnies Freak Out)

## Load Packages
```{r}
library(tidyverse) #v. 1.3.0
library(car) #type 3 sum of squares; v. 3.0.10
library(interactions) #for calculating region-specific slopes; v. 1.1.3
library(rockchalk) #meanCenter; v. 1.8.144
```

## Load data
```{r}
strep_data <- read.csv("data/Love_herbarium_data_n743.csv")
strep_reduced_data <- read.csv("data/Love_herbarium_reduced_dataset_n593.csv")
```

# 1. Temporal Phenological shifts
This code is designed to detect phenological shifts (a) among all specimens and (b) region-specific phenological shifts

### 1a. Phenological shift among all specimens
```{r}
options(contrasts=c("contr.sum", "contr.poly")) #change contrasts so the estimate of the centered variable reflects the mean effect between the two groups (temp groups)

year_model_pooled <- lm(DOY ~ 
                    PI + 
                    year + 
                    elev +
                    latitude +
                    longitude,
                  data = strep_data)

year_model_pooled_center <- meanCenter(year_model_pooled)
summary(year_model_pooled_center)
glance(year_model_pooled_center)

year_model_pooled_ss <- Anova(year_model_pooled_center, type = 3)
year_model_pooled_ss
```

### 1b. Region-specific phenological shifts
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

#reference is cool
year_model <- lm(DOY ~ 
                    PI + 
                    year + 
                    temp_group +
                    year*temp_group +
                    elev +
                    latitude +
                    longitude,
                  data = strep_reduced_data)

year_model_center <- meanCenter(year_model)
summary(year_model_center)
glance(year_model_center)
year_model_ss <- Anova(year_model_centered, type = 3)
year_model_ss
  
sim_slopes(year_model_center, pred = yearc, modx = temp_group, digits = 5)
```

# 2. Phenological sensitivity to climate
### 2a. bFFP
Region-specific sensitivity to the beginning of the frost-free period (day of year of onset) during the year of specimen collection
```{r}
options(contrasts=c("contr.sum", "contr.poly")) 

sens_bFFP <- lm(DOY ~
                     PI +
                     temp_group +
                     bFFP_YOC +
                     MAP_YOC +
                     bFFP_YOC*temp_group,
                   data = strep_reduced_data)

sens_bFFP_centered <- meanCenter(sens_bFFP)
summary(sens_bFFP_centered)

#Anova
sens_bFFP_centered_ss <- Anova(sens_bFFP_centered, type = 3)
sens_bFFP_centered_ss

#simple slopes
sim_slopes(sens_bFFP_centered, pred = bFFP_YOCc, modx = temp_group, digits = 5)

```

### 2b. NFFD
Region-specific sensitivity to the number of frost-free days (frost-free period in days) during the year of specimen collection
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

sens_FFP <- lm(DOY ~
                     PI +
                     temp_group +
                     FFP_YOC +
                     MAP_YOC +
                     FFP_YOC*temp_group,
                   data = strep_reduced_data)

sens_FFP_centered <- meanCenter(sens_FFP)
summary(sens_FFP_centered)

#Anova
sens_FFP_centered_ss <- Anova(sens_FFP_centered, type = 3)
sens_FFP_centered_ss

#simple slopes
sim_slopes(sens_FFP_centered, pred = FFP_YOCc, modx = temp_group, digits = 5)

```

### 2c. MAP
Region-specific sensitivity to mean annual precipitation (mm) during the year of specimen collection
```{r}
options(contrasts=c("contr.sum", "contr.poly")) 

sens_MAP <- lm(DOY ~
                 PI +
                 temp_group +
                 MAP_YOC +
                 MAT_YOC +
                 MAP_YOC*temp_group,
               data = strep_reduced_data)

sens_MAP_centered <- meanCenter(sens_MAP)
summary(sens_MAP_centered)

#Anova
sens_MAP_centered_ss <- Anova(sens_MAP_centered, type = 3)
sens_MAP_centered_ss
```

### 2d. MAT
Region-specific sensitivity to mean annual temperature (°C) during the year of specimen collection
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

sens_MAT <- lm(DOY ~
                     PI +
                     temp_group +
                     MAT_YOC +
                     MAP_YOC +
                     MAT_YOC*temp_group,
                   data = strep_reduced_data)

sens_MAT_centered <- meanCenter(sens_MAT)
summary(sens_MAT_centered)

#Anova
sens_MAT_centered_ss <- Anova(sens_MAT_centered, type = 3)

#simple slopes
sim_slopes(sens_MAT_centered, pred = MAT_YOCc, modx = temp_group, digits = 5)
```

### 2e. PAS
Region-specific sensitivity to precipitation as snow (PAS; mm) during the year of specimen collection
```{r}
options(contrasts=c("contr.sum", "contr.poly")) 

sens_PAS <- lm(DOY ~
                 PI +
                 temp_group +
                 PAS_YOC +
                 MAT_YOC +
                 PAS_YOC*temp_group,
               data = strep_reduced_data)

sens_PAS_centered <- meanCenter(sens_PAS)
summary(sens_PAS_centered)

#Anova
sens_PAS_centered_ss <- Anova(sens_PAS_centered, type = 3)
sens_PAS_centered_ss 

#simple slopes
sim_slopes(sens_PAS_centered, pred = PAS_YOCc, modx = temp_group, digits = 5)
```

### 2f. Average spring/summer Temperature
Sensitivity to spring/summer average temperature (°C) during the year of specimen collection
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

sens_Tave_spsm <- lm(DOY ~
                     PI +
                     temp_group +
                     Tave_sp_sm_YOC +
                     MAP_YOC +
                     Tave_sp_sm_YOC*temp_group,
                   data = strep_reduced_data)

sens_Tave_spsm_centered <- meanCenter(sens_Tave_spsm)
summary(sens_Tave_spsm_centered)

#Anova
sens_Tave_spsm_centered_ss <- Anova(sens_Tave_spsm_centered, type = 3)
sens_Tave_spsm_centered_ss

#simple slopes
sim_slopes(sens_Tave_spsm_centered, pred = Tave_sp_sm_YOCc, modx = temp_group, digits = 5)
```

### 2g. Maximum spring/summer temperature
Region-specific sensitivity to maximum spring/summer temperatures (°C) during the year of specimen collection
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

sens_Tmax_spsm <- lm(DOY ~
                     PI +
                     temp_group +
                     Tmax_sp_sm_ave_YOC +
                     MAP_YOC +
                     Tmax_sp_sm_ave_YOC*temp_group,
                   data = strep_reduced_data)

sens_Tmax_spsm_centered <- meanCenter(sens_Tmax_spsm)
summary(sens_Tmax_spsm_centered)

#Anova
sens_Tmax_spsm_centered_ss <- Anova(sens_Tmax_spsm_centered, type = 3)
sens_Tmax_spsm_centered_ss

#simple slopes
sim_slopes(sens_Tmax_spsm_centered, pred = Tmax_sp_sm_ave_YOCc, modx = temp_group, digits = 5)

```

### 2h. Minimum spring/summer temperature
Region-specific sensitivity to minimum spring/summer temperatures (°C) during the year of specimen collection
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

sens_Tmin_spsm <- lm(DOY ~
                     PI +
                     temp_group +
                     Tmin_sp_sm_ave_YOC +
                     MAP_YOC +
                     Tmin_sp_sm_ave_YOC*temp_group,
                   data = strep_reduced_data)

sens_Tmin_spsm_centered <- meanCenter(sens_Tmin_spsm)
summary(sens_Tmin_spsm_centered)

#Anova
sens_Tmin_spsm_centered_ss <- Anova(sens_Tmin_spsm_centered, type = 3)
sens_Tmin_spsm_centered_ss

#simple slopes
sim_slopes(sens_Tmin_spsm_centered, pred = Tmin_sp_sm_ave_YOCc, modx = temp_group, digits = 5)
```

### 2i. Average winter temperature
Region-specific sensitivity to average winter temperatures (°C) during the year of specimen collection
```{r}
options(contrasts=c("contr.sum", "contr.poly")) 

sens_Tave_wt <- lm(DOY ~
                     PI +
                     temp_group +
                     Tave_wt_YOC +
                     MAP_YOC +
                     Tave_wt_YOC*temp_group,
                   data = strep_reduced_data)

sens_Tave_wt_centered <- meanCenter(sens_Tave_wt)
summary(sens_Tave_wt_centered)

#Anova
sens_Tave_wt_centered_ss <- Anova(sens_Tave_wt_centered, type = 3)
sens_Tave_wt_centered_ss 

#simple slopes
sim_slopes(sens_Tave_wt_centered, pred = Tave_wt_YOCc, modx = temp_group, digits = 5)
```

### 2j. Maximum winter temperature
Region-specific sensitivity to maximum winter temperatures (°C) during the year of specimen collection
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

sens_Tmax_wt <- lm(DOY ~
                     PI +
                     temp_group +
                     Tmax_wt_YOC +
                     MAP_YOC +
                     Tmax_wt_YOC*temp_group,
                   data = strep_reduced_data)

sens_Tmax_wt_centered <- meanCenter(sens_Tmax_wt)
summary(sens_Tmax_wt_centered)

#Anova
sens_Tmax_wt_centered_ss <- Anova(sens_Tmax_wt_centered, type = 3)
sens_Tmax_wt_centered_ss

#simple slopes
sim_slopes(sens_Tmax_wt_centered, pred = Tmax_wt_YOCc, modx = temp_group, digits = 5)
```

### 2k. Minimum winter temperature
Region-specific sensitivity to minimum winter temperatures (°C) during the year of specimen collection
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

sens_Tmin_wt <- lm(DOY ~
                     PI +
                     temp_group +
                     Tmin_wt_YOC +
                     MAP_YOC +
                     Tmin_wt_YOC*temp_group,
                   data = strep_reduced_data)

sens_Tmin_wt_centered <- meanCenter(sens_Tmin_wt)
summary(sens_Tmin_wt_centered)

#Anova
sens_Tmin_wt_centered_ss <- Anova(sens_Tmin_wt_centered, type = 3)
sens_Tmin_wt_centered_ss

#simple slopes
sim_slopes(sens_Tmin_wt_centered, pred = Tmin_wt_YOCc, modx = temp_group, digits = 5)
```

# 3. Temporal change in climate during the collection period
### 3a. bFFP
Region-specific change in the beginning of the frost-free period (day of year)
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

#bFFP
bFFP_lm <- lm(bFFP_YOC ~ 
                      year + 
                      temp_group +
                      elev +
                      latitude +
                      longitude +
                      MAP_YOC +
                      year*temp_group,
                   data = strep_reduced_data)

bFFP_lm_center <- meanCenter(bFFP_lm)
summary(bFFP_lm_center)

#Anova
bFFP_lm_center_ss <- Anova(bFFP_lm_center, type = 3)
bFFP_lm_center_ss

#simple slopes analysis
sim_slopes(bFFP_lm_center, pred = yearc, modx = temp_group, digits = 5)
```

### 3b. NFFD
Region-specific change in the number of frost-free days (frost-free period; days)
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

#FFP
FFP_lm <- lm(FFP_YOC ~ 
                      year + 
                      temp_group +
                      elev +
                      latitude +
                      longitude +
                      MAP_YOC +
                      year*temp_group,
                   data = strep_reduced_data)

FFP_lm_center <- meanCenter(FFP_lm)
summary(FFP_lm_center)


#type 3 SS
FFP_lm_center_ss <- Anova(FFP_lm_center, type = 3)
FFP_lm_center_ss

#simple slopes analysis
sim_slopes(FFP_lm_center, pred = yearc, modx = temp_group, digits = 5)
```

### 3c. MAP
Region-specific changes in MAP (mm)
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

#MAP
MAP_lm <- lm(MAP_YOC ~ 
                      year + 
                      temp_group +
                      elev +
                      latitude +
                      longitude +
                      MAT_YOC +
                      year*temp_group,
                   data = strep_reduced_data)

MAP_lm_center <- meanCenter(MAP_lm)
summary(MAP_lm_center)

#type 3 SS
MAP_lm_center_ss <- Anova(MAP_lm_center, type = 3)
MAP_lm_center_ss 

#simple slopes
sim_slopes(MAP_lm_center, pred = yearc, modx = temp_group, digits = 5)
```

### 3d. MAT
Region-specific change in mean annual temperature (°C) 
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

MAT_lm <- lm(MAT_YOC ~ year + 
               temp_group + 
               elev +
               latitude +
               longitude +
               MAP_YOC +
               year*temp_group, 
             data = strep_reduced_data)

MAT_lm_centered <- meanCenter(MAT_lm)
summary(MAT_lm_centered)

#type 3 SS
MAT_lm_centered_ss <- Anova(MAT_lm_centered, type = 3)
MAT_lm_centered_ss
```

### 3e. PAS
Region-specific changes in precipitation as snow (mm)
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

#PAS
PAS_lm <- lm(PAS_YOC ~ 
                      year + 
                      temp_group +
                      latitude +
                      longitude +
                      elev + 
                      MAT_YOC +
                      year*temp_group,
                   data = strep_reduced_data)

PAS_lm_center <- meanCenter(PAS_lm)
summary(PAS_lm_center)

#type 3 SS
PAS_lm_center_ss <- Anova(PAS_lm_center, type = 3)
PAS_lm_center_ss

#simple slopes analysis
sim_slopes(PAS_lm_center, pred = yearc, modx = temp_group, digits = 5)

```

### 3f. Averate spring/summer temperature
Region-specific changes in mean spring/summer temperature
```{r}
options(contrasts=c("contr.sum", "contr.poly")) 

Tave_sp_sm_lm <- lm(Tave_sp_sm_YOC ~ 
                      year + 
                      temp_group +
                      elev +
                      latitude +
                      longitude +
                      MAP_YOC +
                      year*temp_group,
                   data = strep_reduced_data)

Tave_sp_sm_lm_center <- meanCenter(Tave_sp_sm_lm)
summary(Tave_sp_sm_lm_center)

#type 3 SS
Tave_sp_sm_lm_center_ss <- Anova(Tave_sp_sm_lm_center, type = 3)
Tave_sp_sm_lm_center_ss 
```
### 3g. Spring/summer maximum temperatures
Region-specific changes in spring/summer maximum temperatures
```{r}
options(contrasts=c("contr.sum", "contr.poly")) 

Tmax_sp_sm_lm <- lm(Tmax_sp_sm_ave_YOC ~ 
                      year + 
                      temp_group +
                      elev +
                      latitude +
                      longitude +
                      MAP_YOC +
                      year*temp_group,
                   data = strep_reduced_data)

Tmax_sp_sm_lm_center <- meanCenter(Tmax_sp_sm_lm)
summary(Tmax_sp_sm_lm_center)

#type 3 SS
Tmax_sp_sm_lm_center_ss <- Anova(Tmax_sp_sm_lm_center, type = 3)
Tmax_sp_sm_lm_center_ss
```

### 3h. Spring/summer minimum temperatures
Region-specific changes in spring/summer minimum temperatures
```{r}
options(contrasts=c("contr.sum", "contr.poly")) 

Tmin_sp_sm_lm <- lm(Tmin_sp_sm_ave_YOC ~ 
                      year + 
                      temp_group +
                      elev +
                      latitude +
                      longitude +
                      MAP_YOC +
                      year*temp_group,
                   data = strep_reduced_data)

Tmin_sp_sm_lm_center <- meanCenter(Tmin_sp_sm_lm)
summary(Tmin_sp_sm_lm_center)

#type 3 SS
Tmin_sp_sm_lm_center_ss <- Anova(Tmin_sp_sm_lm_center, type = 3)
Tmin_sp_sm_lm_center_ss

#simple slopes analysis
sim_slopes(Tmin_sp_sm_lm_center, pred = yearc, modx = temp_group, digits = 5)

```

### 3i. Average winter temperatures
```{r}
options(contrasts=c("contr.sum", "contr.poly")) 

Tave_wt_lm <- lm(Tave_wt_YOC ~ 
                      year + 
                      temp_group +
                      elev +
                      latitude +
                      longitude +
                      MAP_YOC +
                      year*temp_group,
                   data = strep_reduced_data)

Tave_wt_lm_center <- meanCenter(Tave_wt_lm)
summary(Tave_wt_lm_center)

#type 3 SS
Tave_wt_lm_center_ss <- Anova(Tave_wt_lm_center, type = 3)
Tave_wt_lm_center_ss
```

### 3j. Spring/summer Maximum temperature
Region-specific change in mean spring/summer temperature (°C)
```{r}
options(contrasts=c("contr.sum", "contr.poly")) 

Tmax_sp_sm_lm <- lm(Tmax_sp_sm_ave_YOC ~ 
                      year + 
                      temp_group +
                      elev +
                      latitude +
                      longitude +
                      MAP_YOC +
                      year*temp_group,
                   data = strep_reduced_data)

Tmax_sp_sm_lm_center <- meanCenter(Tmax_sp_sm_lm)
summary(Tmax_sp_sm_lm_center)

#type 3 SS
Tmax_sp_sm_lm_center_ss <- Anova(Tmax_sp_sm_lm_center, type = 3)
```

###3k. Winter minimum temperature
```{r}
options(contrasts=c("contr.sum", "contr.poly"))

Tmin_wt_lm <- lm(Tmin_wt_YOC ~ 
                      year + 
                      temp_group +
                      elev +
                      latitude +
                      longitude +
                      MAP_YOC +
                      year*temp_group,
                   data = strep_reduced_data)

Tmin_wt_lm_center <- meanCenter(Tmin_wt_lm)
summary(Tmin_wt_lm_center)

#type 3 SS
Tmin_wt_lm_center_ss <- Anova(Tmin_wt_lm_center, type = 3)
Tmin_wt_lm_center_ss 

#simple slopes analysis
sim_slopes(Tmin_wt_lm_center, pred = yearc, modx = temp_group, digits = 5)
```

